### 2023-03-02 chapters/fairness/fif/introduction.tex
minor fix: section -> chapter

	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Section~\ref{fairness_fairXplainer_sec:fifs}). Our method is based on transforming existing fairness metrics into the difference of scaled conditional variances of classifier's prediction, which we then decompose using Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}. FIFs have several desirable properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}), including the decomposability property~\cite{begley2020explainability,lundberg2020explaining}. This property states that the sum of FIFs of all individual and intersectional features equals the bias of the classifier. With this formulation of FIFs, we can identify which features have the greatest influence on bias by looking at the disparity in their scaled decomposed variance between sensitive groups.

%	\item \textbf{Formalism:} Given a classifier, a dataset, and a group fairness metric, we propose to estimate individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier induced by the features (Section~\ref{fairness_fairXplainer_sec:fifs}).	Our FIF formulation is based on representing existing fairness metrics as the difference of the scaled conditional variances of classifier's prediction, followed by a decomposition of variance according to Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}. The presented FIF formulation results in different desired properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}): the efficiency property~\cite{begley2020explainability,lundberg2020explaining} states that the sum of FIFs of all subsets of features is equal to the bias of the classifier.  This allows us to view FIFs as a way to decompose the bias of the classifier among features. In this decomposition, features that demonstrate a higher disparity in the scaled decomposed variance of prediction between sensitive groups are attributed to be highly influential on the bias in prediction.

	\item \textbf{Algorithmic:} We propose a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric. The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}). Building on GSA\cite{saltelli2008global} techniques, {\fairXplainer} solves a local regression problem~\cite{loader2006local} based on cubic splines~\cite{li2010global} to decompose the variance of the classifier's prediction among all the subsets of features.

%	\item \textbf{Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for estimating FIFs of  features for a given classifier, a dataset, and any group fairness metric, namely statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}).  The key idea is to import techniques from GSA\cite{saltelli2008global} to decompose the variance of the classifier's prediction among all  subsets of features by applying a local regression technique~\cite{loader2006local} based on cubic splines~\cite{li2010global}.

	\item \textbf{Experimental:} We evaluate  {\fairXplainer}  on a variety of real-world datasets and machine learning classifiers to demonstrate its efficiency in estimating the individual and intersectional FIFs of features. Our results show that {\fairXplainer} has a higher accuracy in approximating bias using estimated FIFs compared to existing methods (Section~\ref{fairness_fairXplainer_sec:experiments}). Our estimation of FIFs also shows a strong correlation with fairness interventions. Furthermore, {\fairXplainer} yields more granular explanation of the sources of bias by combining both individual and intersectional FIFs, and also detects patterns that existing fairness explainers cannot. Finally, {\fairXplainer} enables us to observe changes in FIFs as a result of different fairness enhancing algorithms~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} and fairness reducing attacks~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning}. This creates opportunities to further examine the impact of these algorithms.	
\end{enumerate}

@@

	Using the methods described in this chapter, we examine the sources of unfairness in DT$ 1 $ by calculating the FIFs of each feature subset. Positive values represent a reduction in fairness due to increased statistical parity, while negative values indicate an improvement in fairness. The results, shown in the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, indicate that fitness $(\mathrm{FIF} = 0.355)$, income $(\mathrm{FIF} = 0.098)$, and the combination of income and fitness $(\mathrm{FIF} = 0.055)$ contribute to higher statistical parity and bias. Fitness, being the root of DT$ 1 $, has a greater impact on statistical parity than income, which is at the second level of DT$ 1 $. Our method also reveals the joint effect of income and fitness on statistical parity, which prior methods do not account for (Section~\ref{fairness_fairXplainer_sec:related_work}). The total of the FIFs of all features, $ 0.355 + 0.098 + 0.055 = 0.507 \approx 0.53 $, approximately matches the statistical parity of the classifier, providing a way to break down the bias among all feature subsets. Note that we discuss the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.

%	Now, applying the techniques developed in this paper, we investigate the sources of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers indicate reduction in fairness by increasing statistical parity, while negative numbers indicate improvement in fairness}. In the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, fitness $(\mathrm{FIF} = 0.35)$, income $(\mathrm{FIF}  = 0.1)$, and the joint effect of income and fitness $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e. higher bias. Intuitively, fitness being at the root of DT$ 1 $ is more influential on statistical parity than income, which is situated at the second level of DT$ 1 $. Our method also reveals the joint (or intersectional) effect of income and fitness on statistical parity, which prior approaches do not consider (Section~\ref{fairness_fairXplainer_sec:related_work}).  Finally, the sum of FIFs of features, $ 0.35 + 0.1 + 0.05 = 0.5 \approx 0.53 $, approximates\footnote{\red{We elaborate on the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.}} the statistical parity of the classifier, thereby demonstrating a way to decompose the bias of the classifier among all the subsets of features.

@@

\section{Related Work}
\label{fairness_fairXplainer_sec:related_work}
Recently, local explanation methods have been applied to black-box classifiers to explain sources of bias through feature-attribution\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition\cite{pan2021explaining}. Our work uses the feature-attribution approach and highlights three limitations of existing methods: (i) a failure to estimate intersectional FIFs, (ii) inaccuracies in approximating bias based on FIFs, and (ii) less correlation of FIFs to fairness intervention. Elaborately, to interpret group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}. They apply a global aggregation (i.e. expectation) of all local explanations corresponding to all data points in a dataset. Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}).  In addition, existing methods ignore the joint contribution of correlated features on bias. To address these limitations, \textit{we develop a formal framework to interpret sources of group unfairness in a classifier and also a novel methodology to estimate FIFs. To the best of our knowledge, this is the first work to do both}.

%Recently, several studies apply \emph{local explanation methods} for black-box classifiers to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}. Our work adopts feature-attribution approach and reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in approximating bias in terms of estimated FIFs and (ii) \emph{failing to estimate intersectional} FIFs. Elaborately, to explain group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}. They apply a global aggregation (such as expectation) of all local explanations corresponding to all data points in a dataset. Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}). In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the bias of the classifier. To this end,  in this paper, \textit{we develop a formal framework to explain sources of group unfairness in a classifier and also a novel methodology to address it. To the best of our knowledge, this is the first work to do both}. %\todo{1. GSA with fairness, 2. Citation not correct, 3. Influence of Yair Zick.}

>>>>

	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Chapter~\ref{fairness_fairXplainer_sec:fifs}). Our method is based on transforming existing fairness metrics into the difference of scaled conditional variances of classifier's prediction, which we then decompose using Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}. FIFs have several desirable properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}), including the decomposability property~\cite{begley2020explainability,lundberg2020explaining}. This property states that the sum of FIFs of all individual and intersectional features equals the bias of the classifier. With this formulation of FIFs, we can identify which features have the greatest influence on bias by looking at the disparity in their scaled decomposed variance between sensitive groups.

%	\item \textbf{Formalism:} Given a classifier, a dataset, and a group fairness metric, we propose to estimate individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier induced by the features (Chapter~\ref{fairness_fairXplainer_sec:fifs}).	Our FIF formulation is based on representing existing fairness metrics as the difference of the scaled conditional variances of classifier's prediction, followed by a decomposition of variance according to Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}. The presented FIF formulation results in different desired properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}): the efficiency property~\cite{begley2020explainability,lundberg2020explaining} states that the sum of FIFs of all subsets of features is equal to the bias of the classifier.  This allows us to view FIFs as a way to decompose the bias of the classifier among features. In this decomposition, features that demonstrate a higher disparity in the scaled decomposed variance of prediction between sensitive groups are attributed to be highly influential on the bias in prediction.

	\item \textbf{Algorithmic:} We propose a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric. The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}). Building on GSA\cite{saltelli2008global} techniques, {\fairXplainer} solves a local regression problem~\cite{loader2006local} based on cubic splines~\cite{li2010global} to decompose the variance of the classifier's prediction among all the subsets of features.

%	\item \textbf{Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for estimating FIFs of  features for a given classifier, a dataset, and any group fairness metric, namely statistical parity, equalized odds, or predictive parity (Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}).  The key idea is to import techniques from GSA\cite{saltelli2008global} to decompose the variance of the classifier's prediction among all  subsets of features by applying a local regression technique~\cite{loader2006local} based on cubic splines~\cite{li2010global}.

	\item \textbf{Experimental:} We evaluate  {\fairXplainer}  on a variety of real-world datasets and machine learning classifiers to demonstrate its efficiency in estimating the individual and intersectional FIFs of features. Our results show that {\fairXplainer} has a higher accuracy in approximating bias using estimated FIFs compared to existing methods (Chapter~\ref{fairness_fairXplainer_sec:experiments}). Our estimation of FIFs also shows a strong correlation with fairness interventions. Furthermore, {\fairXplainer} yields more granular explanation of the sources of bias by combining both individual and intersectional FIFs, and also detects patterns that existing fairness explainers cannot. Finally, {\fairXplainer} enables us to observe changes in FIFs as a result of different fairness enhancing algorithms~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} and fairness reducing attacks~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning}. This creates opportunities to further examine the impact of these algorithms.	
\end{enumerate}

@@

	Using the methods described in this chapter, we examine the sources of unfairness in DT$ 1 $ by calculating the FIFs of each feature subset. Positive values represent a reduction in fairness due to increased statistical parity, while negative values indicate an improvement in fairness. The results, shown in the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, indicate that fitness $(\mathrm{FIF} = 0.355)$, income $(\mathrm{FIF} = 0.098)$, and the combination of income and fitness $(\mathrm{FIF} = 0.055)$ contribute to higher statistical parity and bias. Fitness, being the root of DT$ 1 $, has a greater impact on statistical parity than income, which is at the second level of DT$ 1 $. Our method also reveals the joint effect of income and fitness on statistical parity, which prior methods do not account for (Chapter~\ref{fairness_fairXplainer_sec:related_work}). The total of the FIFs of all features, $ 0.355 + 0.098 + 0.055 = 0.507 \approx 0.53 $, approximately matches the statistical parity of the classifier, providing a way to break down the bias among all feature subsets. Note that we discuss the estimation error of FIFs in Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.

%	Now, applying the techniques developed in this paper, we investigate the sources of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers indicate reduction in fairness by increasing statistical parity, while negative numbers indicate improvement in fairness}. In the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, fitness $(\mathrm{FIF} = 0.35)$, income $(\mathrm{FIF}  = 0.1)$, and the joint effect of income and fitness $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e. higher bias. Intuitively, fitness being at the root of DT$ 1 $ is more influential on statistical parity than income, which is situated at the second level of DT$ 1 $. Our method also reveals the joint (or intersectional) effect of income and fitness on statistical parity, which prior approaches do not consider (Chapter~\ref{fairness_fairXplainer_sec:related_work}).  Finally, the sum of FIFs of features, $ 0.35 + 0.1 + 0.05 = 0.5 \approx 0.53 $, approximates\footnote{\red{We elaborate on the estimation error of FIFs in Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.}} the statistical parity of the classifier, thereby demonstrating a way to decompose the bias of the classifier among all the subsets of features.

@@

\section{Related Work}
\label{fairness_fairXplainer_sec:related_work}
Recently, local explanation methods have been applied to black-box classifiers to explain sources of bias through feature-attribution\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition\cite{pan2021explaining}. Our work uses the feature-attribution approach and highlights three limitations of existing methods: (i) a failure to estimate intersectional FIFs, (ii) inaccuracies in approximating bias based on FIFs, and (ii) less correlation of FIFs to fairness intervention. Elaborately, to interpret group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}. They apply a global aggregation (i.e. expectation) of all local explanations corresponding to all data points in a dataset. Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Chapter~\ref{fairness_fairXplainer_sec:experiments}).  In addition, existing methods ignore the joint contribution of correlated features on bias. To address these limitations, \textit{we develop a formal framework to interpret sources of group unfairness in a classifier and also a novel methodology to estimate FIFs. To the best of our knowledge, this is the first work to do both}.

%Recently, several studies apply \emph{local explanation methods} for black-box classifiers to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}. Our work adopts feature-attribution approach and reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in approximating bias in terms of estimated FIFs and (ii) \emph{failing to estimate intersectional} FIFs. Elaborately, to explain group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}. They apply a global aggregation (such as expectation) of all local explanations corresponding to all data points in a dataset. Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Chapter~\ref{fairness_fairXplainer_sec:experiments}). In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the bias of the classifier. To this end,  in this paper, \textit{we develop a formal framework to explain sources of group unfairness in a classifier and also a novel methodology to address it. To the best of our knowledge, this is the first work to do both}. %\todo{1. GSA with fairness, 2. Citation not correct, 3. Influence of Yair Zick.}

### 2023-03-02 chapters/fairness/fvgm/framework.tex
minor fix: section -> chapter

\paragraph{A Heuristic for Faster Computation.} 
We propose two improvements for a faster computation of the dynamic programming solution. Firstly, we observe that in Eq.~\eqref{fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights. Hence, we reorder $ \mathbf{B} $ such that existential/universal variables appear earlier in $ \mathbf{B} $ than random variables. This allows us to avoid unnecessary repeated exploration of existential/universal variables in $ \mathsf{dp} $. Moreover, according to the remark in Section~\ref{fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}. Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables. In particular, if $ \tau \le 0.5(W_{pos} - W_{neg}) $, $ \tau $ is closer to $ W_{pos} $ than $ W_{neg} $. Hence, we sort each cluster in descending order of weights. Otherwise,  we sort in ascending order. We illustrate our dynamic programming approach in Example~\ref{fvgm_example:subset-sum}.

>>>>

\paragraph{A Heuristic for Faster Computation.} 
We propose two improvements for a faster computation of the dynamic programming solution. Firstly, we observe that in Eq.~\eqref{fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights. Hence, we reorder $ \mathbf{B} $ such that existential/universal variables appear earlier in $ \mathbf{B} $ than random variables. This allows us to avoid unnecessary repeated exploration of existential/universal variables in $ \mathsf{dp} $. Moreover, according to the remark in Chapter~\ref{fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}. Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables. In particular, if $ \tau \le 0.5(W_{pos} - W_{neg}) $, $ \tau $ is closer to $ W_{pos} $ than $ W_{neg} $. Hence, we sort each cluster in descending order of weights. Otherwise,  we sort in ascending order. We illustrate our dynamic programming approach in Example~\ref{fvgm_example:subset-sum}.

### 2023-03-02 chapters/fairness/fvgm/framework2.tex
with new formatting
minor fix: section -> chapter

\input{chapters/fairness/fvgm/dp_illustration}	
\subsection{{\stochastic} with Correlated Variables} 
\label{fvgm_sec:dp_with_BN}
In {\stochastic} presented in Section~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent. This independence assumption often leads to an \textit{inaccurate estimate} of the probability of positive prediction of the classifier because both sensitive and non-sensitive features can be correlated in practical fairness problems. Therefore, we extend {\stochastic} to include correlations among variables.

We consider a Bayesian network $ \BN = (\graph, \theta) $ to represent correlated variables, where $ G \triangleq (\mathbf{V}, \mathbf{E}) $, $ \mathbf{V} \subseteq \mathbf{B} $, $ \mathbf{E} \subseteq \mathbf{V} \times \mathbf{V}  $, and $ \theta $ is the parameter of the network.  In  $ \BN $, we constrain that there is no conditional probability of choice (i.e., existential and universal) variables as we optimize their assignment in {\stochastic}. Choice variables, however, can impose conditions on chance (i.e., random) variables. In practice, we achieve this by allowing no incoming edge on choice variables while learning $ \BN $ (ref. Section~\ref{fvgm_sec:experiments}).

@@

Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to enumerate the assignment of parents before computing $ \mathsf{dp} $ on the child. Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}) $ for $ \mathbf{u} \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities into $ \stochastic $.  For this enumeration, we do not store $ \mathsf{dp}(i, \tau, \mathbf{u}) $ in memory. However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Section~\ref{fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.  Lemma~\ref{fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).

@@

\begin{proof}
	We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $. For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{fvgm_sec:dp_with_BN}.

@@

	\begin{example}
		\normalfont
		We extend Example~\ref{fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $. Parameters $ \theta $ imply conditional probabilities $ \Pr[Q|P] = 0.6 $ and $ \Pr[Q|\neg P] = 0.3 $. 	In Figure~\ref{fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $. We, however, observe that the dynamic programming solution in Section~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $. Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum probability of positive prediction of the classifier as $ 0.65 $ for $ P = 1 $. The minimum probability of positive prediction (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $. 
	\end{example}

>>>>

\input{chapters/fairness/fvgm/dp_illustration}	
\subsection{Stochastic Subset Sum Problem with Correlated Variables} 
\label{fvgm_sec:dp_with_BN}
In {\stochastic} presented in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent. This independence assumption often leads to an \textit{inaccurate estimate} of the probability of positive prediction of the classifier because both sensitive and non-sensitive features can be correlated in practical fairness problems. Therefore, we extend {\stochastic} to include correlations among variables.

We consider a Bayesian network $ \BN = (\graph, \theta) $ to represent correlated variables, where $ G \triangleq (\mathbf{V}, \mathbf{E}) $, $ \mathbf{V} \subseteq \mathbf{B} $, $ \mathbf{E} \subseteq \mathbf{V} \times \mathbf{V}  $, and $ \theta $ is the parameter of the network.  In  $ \BN $, we constrain that there is no conditional probability of choice (i.e., existential and universal) variables as we optimize their assignment in {\stochastic}. Choice variables, however, can impose conditions on chance (i.e., random) variables. In practice, we achieve this by allowing no incoming edge on choice variables while learning $ \BN $ (ref. Chapter~\ref{fvgm_sec:experiments}).

@@

Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to enumerate the assignment of parents before computing $ \mathsf{dp} $ on the child. Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}) $ for $ \mathbf{u} \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities into $ \stochastic $.  For this enumeration, we do not store $ \mathsf{dp}(i, \tau, \mathbf{u}) $ in memory. However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Chapter~\ref{fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.  Lemma~\ref{fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).

@@

\begin{proof}
	We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $. For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Chapter~\ref{fvgm_sec:dp_with_BN}.

@@

	\begin{example}
		\normalfont
		We extend Example~\ref{fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $. Parameters $ \theta $ imply conditional probabilities $ \Pr[Q|P] = 0.6 $ and $ \Pr[Q|\neg P] = 0.3 $. 	In Figure~\ref{fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $. We, however, observe that the dynamic programming solution in Chapter~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $. Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum probability of positive prediction of the classifier as $ 0.65 $ for $ P = 1 $. The minimum probability of positive prediction (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $. 
	\end{example}

### 2023-03-02 chapters/fairness/justicia/introduction.tex
minor fix: section -> chapter

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
In this chapter, we study formal verification of fairness in machine learning, where we verify the bias of a classifier given the probability distribution of features. To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

@@

\begin{example}\label{fairness_justicia_example:intro}
	\normalfont
	Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60). Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}). We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example_distribution}. Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref. Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age. However, a fairness verifier, such as {\justicia}, would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}). This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
\end{example}

>>>>

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
In this chapter, we study formal fairness verification problem in machine learning, where we verify the bias of a classifier given the probability distribution of features. To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

@@

\begin{example}\label{fairness_justicia_example:intro}
	\normalfont
	Let us consider a classification problem (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60). Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}). We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example_distribution}. Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref. Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age. However, a fairness verifier, such as {\justicia}, would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}). This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
\end{example}
